A Mechanised Library for Generic (Co)Programming

### 2024-02-02 paper/paper.tex
minor
Paper

Our goals are: (1) that the mechanisation of hylomorphisms and their laws are
axiom-free; and (2) Coq hylomorphisms can be extracted to recursive OCaml
functions that use no instances of \mintinline{OCaml}{Obj.magic}. Note that
although the use of \mintinline{OCaml}{Obj.magic} is safe in the code extracted
by Coq, its presence may prevent some optimisations by the OCaml
compiler\footnote{\dcas{I read this somewhere, search \& add reference/link}}.

@@

\dcas{A first approach (problems with extraction, lots of Obj.magic, etc)}
\begin{minted}{Coq}
  Record Container (Shape : Set) (Pos : Shape -> Set) (X : Set) :=
    { shape : Shape; contents : Pos shape -> X }.
\end{minted}

\subsection{A Type for Morphisms}

>>>>

Our goals are: (1) the mechanisation of hylomorphisms and their laws should be
axiom-free; and (2) Coq hylomorphisms should be extracted to recursive OCaml
functions that use no instances of \mintinline{OCaml}{Obj.magic}. Note that
although the use of \mintinline{OCaml}{Obj.magic} is safe in the code extracted
by Coq, it increases the Trusted Code Base, makes code less legible for
programmers, and its presence will likely inhibit optimisations by the OCaml
compiler\footnote{\dcas{I read this somewhere, search \& add reference/link?}}.

@@

Mechanising functors in Coq is straightforward. But taking least/greatest
fixpoints of the said functors may fail, due to not being strictly positive. A
common approach to construct strictly positive functors in Coq is to use
\emph{containers}. In essence, a container 
\dcas{A first approach (problems with extraction, lots of Obj.magic, etc)}
\begin{minted}{Coq}
  Record Container (Shape : Set) (Pos : Shape -> Set) (X : Set) :=
    { shape : Shape; contents : Pos shape -> X }.
\end{minted}

\dcas{A better approach for extraction. Unfortunately, to remain axiom-free and
heterogeneous equality we need to have a decidable ``Valid''}
\begin{minted}{Coq}
  Record Container (Shape : Set) (Pos : Set) 
                   (Valid : Shape -> Pos -> Prop) (X : Set) :=
    { shape : Shape; contents : {p | Valid shape pos} -> X }.
\end{minted}

\dcas{Solution to the previous problem}
\begin{minted}{Coq}
  Record Container (Shape : Set) (Pos : Set) 
                   (valid : Shape -> Pos -> bool) (X : Set) :=
    { shape : Shape; contents : {p | valid shape pos = true} -> X }.
\end{minted}

\dcas{A shape, a position, and a decidable predicate on the validity of a
position in a shape define a functor}

\begin{minted}{Coq}
  Coercion is_true (b : bool) : Prop := b = true.
  Class functor (Shape Pos : Set) := { valid : Shape -> Pos -> bool }.
  Record Container `{functor Shape Pos} (X : Set) :=
    { shape : Shape; contents : {p | valid shape pos} -> X }.
\end{minted}

\subsection{Types and Morphisms}

Reasoning about the equality of containers requires reasoning about the
equality of the contained elements. But the contained elements are given by a
function. This means that container equality needs to be extensional, if we
want to avoid the functional extensionality axiom. Furthermore, the equality of
functions also needs to be extensional. We capture this fact by associating
each type with an equivalence relation:

\begin{minted}{Coq}
  Class Equiv (T : Type) := 
  { equiv : T -> T -> Prop;
    e_refl : forall x, equiv x x;
    e_sym : forall x y, equiv x y -> equiv y x;
    e_trans : forall x y z, equiv x y -> equiv y z -> equiv x z;
  }.
\end{minted}

\dcas{Morphisms must respect equivalences}:

\begin{minted}{Coq}
  Record Morph `{Equiv A} `{Equiv B} := 
  { app : A -> B;
    app_ext : forall x y, equiv x y -> equiv (app x) (app y);
  }.

\dcas{We provide common morphisms for polynomials: sums, products, const,
  exponential}

\end{minted}
